Supporting Information:\\Black-swan events in animal populations

### 2016-06-23 text/ms.tex
Incorporate Nick and Trevor 's comments
Also fix formatting of captions

Keywords:
Bayesian,
die-offs,
ecological surprises,
Gompertz,
heavy tails,
leptokurtic,
mass mortality,
population dynamics,
risk,
Student-t distribution

@@

Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
and manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases
and tend to be driven by extrinsic perturbations
such as climate, severe winters, or disease.
Thus, ignoring black-swan events will
underestimate extinction risk.
% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to such extremes.

@@

Indeed, such die-offs may be the most important element
affecting population persistence \citep{mangel1994} and their importance is
likely to increase given projected increases in the frequency and magnitude of
climate-related extremes \citep{ipcc2012}.
% Despite this anecdotal evidence for
% ecological black swans, systematic searches of large numbers of timeseries have
% yet to reveal black-swan events \citep{keitt1998, allen2001, halley2002}.
However, the overwhelming majority of population model fitting and risk forecasting
assumes that deviations from model predictions can be represented by a normal
distribution \citep[on a log scale; e.g.][]{brook2006a, knape2012}.
If black swans occur, though, a normal distribution would under-estimate the
probability of extreme events occurring.

There are many examples of black-swan events in populations
but we lack a large-scale comparative analysis
with a method that allows for
population dynamics
(e.g.\ per capita productivity, density dependence, and typical stochasticity).
Previous comparative analyses fitting
heavy-tailed distributions
(distributions with higher probabilities of extreme events
than the normal distribution)
to timeseries
have not accounted for an underlying population dynamics model
\citep{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
jumps in abundance that exceed a threshold of decline
\citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
But this approach gives no indication of relative event frequency
and again does not distinguish events from
expected dynamics.
Methodologically, the generalized extreme value (GEV) distribution represents
the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
because the GEV requires maxima per time block
or peaks beyond a chosen threshold,
the GEV requires longer timeseries then typically available
for wild populations and is not simply integrated
into a population dynamics model.

@@

We fit population dynamics models in which the process noise is
drawn from a Student-t distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
% For example, at \(\nu = 2\), an event
% more than 4 standard deviations below the mean is expected
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a t distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution
% more than four standard deviations below the mean is almost 70,000 times higher
% than the probability of drawing such a value from a normal distribution
% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

We picked the t distribution for pragmatic and mechanistic reasons.
Pragmatically, the t distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations are not present.
Mechanistically, the t distribution can be formed from a mixture of normal distributions
with the same mean and different variances \citep{gelman2014}.
This could represent the effect of unmodeled spatial heterogeneity,
nonstationary environmental processes, or
varying responses to temperature among individuals, for example.

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD. The Gompertz model represents population growth as a linear function in log space. If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
\end{align}
The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-t distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails and are therefore evidence of
black-swan events (Fig.~1A, B).

For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our Half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain
convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by \citet{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid drifting towards infinity. In
any case, for \(\nu > 20\) the t distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the t distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the
population as heavy tailed.

@@

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are mostly concerned about when categorizing
populations as heavy tailed. We furthermore considered a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here (not shown).

@@

\subsubsection{Observation error} Observation error can bias parameter
estimates \citep{knape2012} and is known to affect the ability to detect extreme
events \citep{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth
rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed
abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
population growth rate with increases in abundance. In contrast, the Gompertz
model assumes a linear decrease in population growth rate with increases in
\textit{log} abundance.

@@

\subsection{Simulation testing}
We performed two types of simulation testing.
First, we tested how easily the Student-t distribution \(\nu\) parameter could
be recovered given different true values of \(\nu\) and different sample sizes.
Second, we tested the ability of the heavy-tailed Gompertz model to obtain
unbiased parameter estimates of \(\nu\) given that a set of process deviations
was provided in which the effective \(\nu\) value was close to the true \(\nu\)
value.

@@

Our simulation-testing results indicate that the Gompertz model can recapture
the true value of \(\nu\) when the process noise was chosen so that appropriate
tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
of observation error caused the model to tend to underestimate the degree of
heavy-tailedness. Fitting a model with correctly specified observation error
did not make substantial improvements to model bias (Fig.~S\ref{fig:sim-prob}).

@@

\subsection{Skewed Student-t forecasts} To evaluate the apparent skewness
of heavy-tailed process noise we fit Gompertz models with skewed Student-t
distributed process noise \citep{fernandez1998}. This distribution adds one
parameter to the Student-t distribution, \(\gamma\), which controls the
skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
parameter of each model at three levels of evidence for heavy-tails:
\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
(Fig.~4).

@@

We found evidence that black-swan dynamics are infrequent but highly
influential: they occur most frequently for birds (7\%), mammals (5\%), and
insects (3\%) and almost never in fishes for this dataset (Fig.~2). Black-swan
events were taxonomically widespread, occurring in 38\% of taxonomic orders.
Accounting for time series length and partially pooling inference across
taxonomic class and order with a hierarchical model, we found stronger evidence
for black swans in insect populations than these statistics suggest---four of
eight orders with the highest median probability of heavy tails were insect
orders (Fig.~3a).

The majority of our heavy-tailed estimates were robust to alternative
population models, observation error, and choice of Bayesian priors. Our
conclusions were not systematically altered when we included an autocorrelation
structure in the residuals, modelled population growth rates with or without
density dependence, or modelled the population dynamics as Ricker-logistic
instead of Gompertz (Fig.~S\ref{fig:alt}). Similarly, introducing moderate
observation error (CV = 0.2) only slightly decreased the estimated prevalence
of black-swan events (Fig.~S\ref{fig:alt}), and the strength of the prior on
\(\nu\) had little influence on estimates of black-swan dynamics
(Fig.~S\ref{fig:alt-priors}). Finally, our simulation testing shows that, if
anything, our models underpredict the true magnitude and probability of
heavy-tailed events---especially given the time series have a median of only 26
years in our analysis (Figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).

@@

The majority of black-swan events (\percBSDown \%) were downward (die-offs)
rather than upward (unexpectedly rapid abundance increases). Of the black-swan
events with published explanations (Table~S2), the majority involved
a combination of multiple factors. For example, a synchronization of
environmental- and predation-mediated population cycles is thought to have
caused a downward black-swan event for a water vole (\emph{Arvicola
terrestris}) population \citep{saucy1994}. Other black swans were the result of
a sequence of extreme climate events on their own. For instance, severe winters
in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings
in grey heron (\emph{Ardea cinerea}) abundance in the United
Kingdom \citep{stafford1971} (Fig.~4c). Our analysis finds that the last event
was a combination of two heavy-tailed events in a row and the population took
three times longer to recover than predicted \citep{stafford1971}. Downward black
swans were sometimes followed by upward black swans. For example, during
a period of population crowding and nest shortages, a population of European
shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United
Kingdom declined suddenly following a red tide event in 1968 \citep{potts1980}.
This freed up quality nest sites for first-time breeders, productivity
increased, and the population experienced a rapid upswing in
abundance \citep{potts1980}.

@@

We systematically evaluated the presence of black-swan events in hundreds of
animal populations. We find strong evidence for their occurrence. Black-swan
population crashes are substantially more frequent
than black-swan population
increases and are usually driven by external events such as weather and
disease.
Thus our analysis provides strong evidence
for downward-skewed heavy-tailed events in abundance time series of higher
taxa, and ignoring these events will tend to underestimate the risk of
population declines.
\textit{TODO: do we want to have a recap paragraph at the beginning of the discussion now?}

@@

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a t distribution with heavy
tails \citep{gelman2014}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \citep{schindler2010}, or an intrinsic change in population variability
through time \citep{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \citep{nunez2012} including extreme climate
events \citep{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black-swan events, or sharp changes in human
pressures such as fishing or habitat destruction. Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \citep{kirby2009} or a rare alignment of
non-synergistic forces \citep{denny2009}.

There are a number of caveats when considering the generality of our results.
The GPDD data represent a taxonomically and geographically biased sample of
populations---the longer time series we focus on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Additionally, some apparent black-swan events could be
recording mistakes, or conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors. Finally, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.

>>>>

Keywords:
Bayesian,
die-offs,
ecological surprises,
Gompertz,
heavy tails,
leptokurtic,
mass mortality,
population dynamics,
risk,
Student-$t$ distribution

@@

Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
and tend to be driven by extrinsic perturbations
such as climate, severe winters, or disease.
Thus, ignoring black-swan events will
underestimate extinction risk.
% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to ecological surprises.

@@

Indeed, such die-offs may be the most important element
affecting population persistence \citep{mangel1994} and their importance is
likely to increase given projected increases in the frequency and magnitude of
climate-related extremes \citep{ipcc2012}.
% Despite this anecdotal evidence for
% ecological black swans, systematic searches of large numbers of timeseries have
% yet to reveal black-swan events \citep{keitt1998, allen2001, halley2002}.
However, the overwhelming majority of population model fitting and risk forecasting
assumes that deviations from model predictions can be represented by a normal
distribution \citep[on a log scale; e.g.][]{brook2006a, knape2012}.
If black swans occur, though, a normal distribution would under-estimate the
probability of extreme events occurring \citep{taleb2007}.

Whereas there are many reports of black-swan events,
only a flexible comparative approach consistently applied 
to a large number of timeseries can yield insights into the frequency, strength,
and correlates of black-swan events. 
We are unaware of such a comparative analysis.
Previous comparative analyses fitting
heavy-tailed distributions
(distributions with higher probabilities of extreme events
than the normal distribution)
to timeseries
have not accounted for an underlying population dynamics model
\citep{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
sudden changes in abundance that exceed a threshold of decline
\citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
But this approach gives no indication of relative event frequency
and again does not distinguish events from
expected dynamics.
Methodologically, the generalized extreme value (GEV) distribution represents
the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
the GEV requires maxima per time block
or peaks beyond a chosen threshold,
and hence requires longer timeseries then typically available
for wild populations.
Furthermore, the GEV approach cannot be easily integrated into population dynamics models.

@@

We fit population dynamics models in which the process noise is
drawn from a Student-$t$ distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
% For example, at \(\nu = 2\), an event
% more than 4 standard deviations below the mean is expected
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution
% more than four standard deviations below the mean is almost 70,000 times higher
% than the probability of drawing such a value from a normal distribution
% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

We chose the $t$ distribution for a number of reasons. 
Pragmatically, the $t$ distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations are not present.
Furthermore, the $t$ distribution arises from a mixture of normal distributions
with the same mean and different variances \citep{gelman2014},
which mimics the underlying basis for black-swan events.
There are a wide range of mechanistic processes 
that could give rise to extreme events,
such as spatial heterogeneity, nonstationary or heavy-tailed environmental processes, 
or varying responses to temperature among individuals. 
While evaluating these various hypotheses 
would be important for diagnosing the underlying causes 
of extreme events and to develop predictive models, 
it would be impossible to tease out each of these 
underlying causes from population trend data alone in a comparative manner. 
Instead we seek a simple flexible phenomenological model 
that can be applied across the widest possible 
range of population time series to provide a 
comparative overview of the prevalence 
of black-swan events.

This could represent the effect of unmodeled spatial heterogeneity,
nonstationary environmental processes, or
varying responses to temperature among individuals, for example.

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD. The Gompertz model represents population growth as a linear function in log space. If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
\end{align}
The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-$t$ distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails and are therefore evidence of
black-swan events (Fig.~1A, B).

For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by \citet{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
population as heavy tailed.

@@

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are mostly concerned about when categorizing
populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here,
and the results are therefore not shown.

@@

\subsubsection{Observation error} Observation error can bias parameter
estimates \citep{knape2012} and is known to affect the ability to detect extreme
events \citep{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth
rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed
abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
population growth rate with increases in abundance. In contrast, the Gompertz
model assumes a linear decrease in population growth rate with increases in
\textit{ln} abundance.

@@

\subsection{Simulation testing}
We performed two types of simulation testing.
First, we tested how easily the Student-$t$ distribution \(\nu\) parameter could
be recovered given different true values of \(\nu\) and different sample sizes.
Second, we tested the ability of the heavy-tailed Gompertz model to obtain
unbiased parameter estimates of \(\nu\) given that a set of process deviations
was provided in which the effective \(\nu\) value was close to the true \(\nu\)
value.

@@

Our simulation-testing results indicate that the Gompertz model can recapture
the true value of \(\nu\) when the process noise was chosen so that appropriate
tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
of observation error caused the model to tend to underestimate the degree of
heavy-tailedness. Fitting a model with correctly specified observation error
did not substantially reduce model bias (Fig.~S\ref{fig:sim-prob}).

@@

\subsection{Skewed Student-t forecasts} To evaluate the apparent skewness
of heavy-tailed process noise we fit Gompertz models with skewed Student-t
distributed process noise \citep{fernandez1998}. This distribution adds one
parameter to the Student-$t$ distribution, \(\gamma\), which controls the
skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
parameter of each model at three levels of evidence for heavy-tails:
\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
(Fig.~4).

@@

We found evidence that black-swan dynamics are infrequent but highly
influential: in our dataset they were detected most 
frequently for birds (7\%), mammals (5\%), and
insects (3\%) but almost never in fishes (Fig.~2). Black-swan
events were taxonomically widespread, occurring in 38\% of taxonomic orders.
Accounting for time series length and partially pooling inference across
taxonomic class and order with a hierarchical model, we found stronger evidence
for black swans in insect populations than these statistics suggest---four of
eight orders with the highest median probability of heavy tails were insect
orders (Fig.~3a).

The majority of our heavy-tailed estimates were robust to alternative
population models, observation error, and choice of Bayesian priors. Our
conclusions were not systematically altered when we included autocorrelation
in the residuals, modelled population growth rates with or without
density dependence, or modelled the population dynamics as Ricker-logistic
instead of Gompertz (Fig.~S\ref{fig:alt}). Similarly, introducing moderate
observation error (CV = 0.2) only slightly decreased the estimated prevalence
of black-swan events (Fig.~S\ref{fig:alt}), and the strength of the prior on
\(\nu\) had little influence on estimates of black-swan dynamics
(Fig.~S\ref{fig:alt-priors}). Finally, our simulation testing shows that, if
anything, our models underpredict the true magnitude and probability of
heavy-tailed events---especially given that the time series are relatively short,
with a median of only 26
years in our analysis (Figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).

@@

The majority of black-swan events (\percBSDown \%) were downward (die-offs)
rather than upward (unexpectedly rapid abundance increases). Of the black-swan
events with published explanations (Table~S2), the majority involved
a combination of multiple factors. For example, a synchronization of
environmental- and predation-mediated population cycles is thought to have
caused a downward black-swan event for a water vole (\emph{Arvicola
terrestris}) population \citep{saucy1994}. Other black swans were the result of
a sequence of extreme climate events on their own. For instance, severe winters
in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings
in grey herons (\emph{Ardea cinerea}) in the United
Kingdom \citep{stafford1971} (Fig.~4c). Our analysis finds that the last event
was a combination of two heavy-tailed events in a row and the population took
three times longer to recover than predicted \citep{stafford1971}. Downward black
swans were sometimes followed by upward black swans. For example, during
a period of population crowding and nest shortages, a population of European
shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United
Kingdom declined suddenly following a red tide event in 1968 \citep{potts1980}.
This freed up quality nest sites for first-time breeders, productivity
increased, and the population experienced a rapid upswing in
abundance \citep{potts1980}.

@@

We systematically evaluated the presence of black-swan events in hundreds of
animal populations in a consistent manner. 
We find strong evidence for their occurrence in animal populations. Black-swan
population crashes are substantially more frequent
than black-swan population
increases and are usually driven by external events such as weather and
disease.
Thus our analysis provides strong evidence
for downward-skewed heavy-tailed events in abundance time series of higher
taxa, and ignoring these events will tend to underestimate the risk of
population declines.
Next we consider 
(1) the possible mechanisms underlying black-swan events, 
(2) caveats to our findings, 
(3) the consequences of our findings for making risk forecasts, and 
(4) for managing natural resources.

@@

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a $t$ distribution with heavy
tails \citep{gelman2014}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \citep{schindler2010}, or an intrinsic change in population variability
through time \citep{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \citep{nunez2012} including extreme climate
events \citep{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black-swan events, or sharp changes in human
pressures such as fishing or habitat destruction. Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \citep{kirby2009} or a rare alignment of
non-synergistic forces \citep{denny2009}.

There are a number of caveats when considering the generality of our results.
The GPDD data represent a taxonomically and geographically biased sample of
populations---the longer time series we focus on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Additionally, some apparent black-swan events could be
recording mistakes, although conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors. Finally, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.
